---
title: "Greenhouse Gas Emission Analysis"
author: "Pushya Shree Konasale Jayaramu"
email: pushyashree.kj.2000@gmail.com
output:html_document: default
---

## **Introduction**
This report analyzes greenhouse gas emissions for different countries from 1990 to 2014.  
It includes:
- Emission distributions
- Yearly trends
- European vs. Non-European comparisons
- Correlations
- Percentage changes over time  

The dataset is obtained from [UNData](http://data.un.org/Explorer.aspx).

License:
Per the [UNData](http://data.un.org/Explorer.aspx) terms of use: all data and metadata provided on [UNData's](http://data.un.org/Explorer.aspx) website are available free of charge and may be copied freely, duplicated and further distributed provided that [UNData](http://data.un.org/Explorer.aspx) is cited as the reference.

---

## **1️⃣ Data Loading & Preprocessing**
```{r setup, echo=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(ggpubr)
library(ggcorrplot)
library(RColorBrewer)

# Load dataset
ghg_data <- read.csv("greenhouse_gas_inventory_data_data.csv")

# Display first few rows
head(ghg_data)
str(ghg_data)
```

## **2️⃣ Renaming Categories**
```{r, results="hide"}
# Rename emission categories for readability
ghg_data <- ghg_data %>%
  mutate(category = recode(category,
                                    "carbon_dioxide_co2_emissions_without_land_use_land_use_change_and_forestry_lulucf_in_kilotonne_co2_equivalent" = "CO2 (No LULUCF)",
                                    "greenhouse_gas_ghgs_emissions_including_indirect_co2_without_lulucf_in_kilotonne_co2_equivalent" = "GHG (Total, No LULUCF)",
                                    "greenhouse_gas_ghgs_emissions_without_land_use_land_use_change_and_forestry_lulucf_in_kilotonne_co2_equivalent" = "GHG (No LULUCF)",
                                    "hydrofluorocarbons_hfcs_emissions_in_kilotonne_co2_equivalent" = "HFCs",
                                    "methane_ch4_emissions_without_land_use_land_use_change_and_forestry_lulucf_in_kilotonne_co2_equivalent" = "CH4 (No LULUCF)",
                                    "nitrogen_trifluoride_nf3_emissions_in_kilotonne_co2_equivalent" = "NF3",
                                    "nitrous_oxide_n2o_emissions_without_land_use_land_use_change_and_forestry_lulucf_in_kilotonne_co2_equivalent" = "N2O (No LULUCF)",
                                    "perfluorocarbons_pfcs_emissions_in_kilotonne_co2_equivalent" = "PFCs",
                                    "sulphur_hexafluoride_sf6_emissions_in_kilotonne_co2_equivalent" = "SF6",
                                    "unspecified_mix_of_hydrofluorocarbons_hfcs_and_perfluorocarbons_pfcs_emissions_in_kilotonne_co2_equivalent" = "HFCs + PFCs (Mixed)"
))
```

## **3️⃣ Assigning Regions (Europe vs. Non-Europe)**
```{r, results="hide"}
# Define European countries
european_countries <- c("Austria", "Belarus", "Belgium", "Croatia", "Czech Republic", "Denmark",
                        "Estonia", "European Union", "Finland", "France", "Germany", "Greece",
                        "Hungary", "Iceland", "Ireland", "Italy", "Latvia", "Liechtenstein",
                        "Lithuania", "Luxembourg", "Malta", "Monaco", "Netherlands", "Norway",
                        "Poland", "Romania", "Slovakia", "Slovenia", "Spain", "Sweden", "Switzerland")

# Assign region
ghg_data <- ghg_data %>%
  mutate(region = ifelse(country_or_area %in% european_countries, "Europe", "Non-Europe"))
```

## **4️⃣ Visualizing Emission Distributions: Emission Distribution in 1990 vs 2014**
```{r, fig.width=10, fig.height=6}
filtered_data <- ghg_data %>% filter(year %in% c(1990, 2014))

ggplot(filtered_data, aes(x = as.factor(year), y = value, fill = category)) +
  geom_boxplot(outlier.shape = NA) + 
  scale_y_log10() +  
  facet_wrap(~ category, scales = "free_y") +
  labs(title = "Emission Distribution in 1990 vs 2014 (All Countries)",
       x = "Year", y = "Emission Value (log scale)") +
  theme_minimal()
```

## **5️⃣ European vs Non-European Emission Comparison**
```{r, fig.width=10, fig.height=6}
ggplot(ghg_data, aes(x = region, y = value, fill = category)) +
  geom_boxplot(outlier.shape = NA) +
  scale_y_log10() +  
  labs(title = "Comparison of Emissions: Europe vs Non-Europe",
       x = "Region", y = "Emission Value (log scale)") +
  theme_minimal()
```

## **6️⃣ How Emission Categories Have Changed Over Time (1990-2014)**
```{r, fig.width=10, fig.height=6}
ggplot(ghg_data, aes(x = year, y = value, color = category)) +
  geom_line(stat="summary", fun="mean", linewidth=1.2) + 
  labs(title = "Yearly Trends of Different Emission Categories",
       x = "Year", y = "Mean Emission Value") +
  theme_minimal()
```

## **7️⃣ Comparing Emission Distributions Between 1990 and 2014**
```{r, fig.width=10, fig.height=6}
filtered_data <- ghg_data %>% filter(year %in% c(1990, 2014))

ggplot(filtered_data, aes(x = value, fill = as.factor(year))) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ category, scales = "free") +
  scale_x_log10() +
  labs(title = "Emission Distribution Comparison (1990 vs 2014)",
       x = "Emission Value (log scale)", y = "Density") +
  theme_minimal()
```

## **8️⃣ Understanding Relationships Between Different Greenhouse Gases**
```{r, fig.width=10, fig.height=6}
# Reshape data to wide format for correlation
wide_data <- ghg_data %>%
  select(country_or_area, year, category, value) %>%
  pivot_wider(names_from = category, values_from = value)

# Compute correlation matrix
emission_corr <- cor(wide_data[, -c(1,2)], use = "complete.obs")

ggcorrplot(emission_corr, method = "square", type = "lower", lab = TRUE, colors= c("#d8b365", "#f5f5f5", "#5ab4ac") ) +
  ggtitle("Correlation Between Emission Categories")

```


```{r}
# Aggregate emissions for each country and category
agg_ghg_data <- ghg_data %>%
  group_by(country_or_area, category, year) %>%
  summarize(total_value = sum(value, na.rm = TRUE), .groups = 'drop') %>%
  filter(year %in% c(1990, 2014))  # Keep only 1990 and 2014

# Reshape data to wide format for 1990 vs 2014
change_data <- agg_ghg_data %>%
  pivot_wider(names_from = year, values_from = total_value, names_prefix = "year_") %>%
  mutate(Percentage_Change = ((year_2014 - year_1990) / year_1990) * 100) %>%
  filter(!is.na(Percentage_Change))  # Remove missing values

change_data <- change_data %>%
  mutate(Percentage_Change = pmin(pmax(Percentage_Change, 0), 100))

library(RColorBrewer)
# Create a green-to-red color ramp
color_palette <- colorRampPalette(rev(brewer.pal(9, "RdYlGn")))(100)

# Assign region classification
change_data <- change_data %>%
  mutate(region = ifelse(country_or_area %in% european_countries, "Europe", "Non-Europe"))

# Function to plot heatmap
plot_heatmap <- function(region_filter) {
  ggplot(change_data %>% filter(region == region_filter),
         aes(x = category, y = reorder(country_or_area, Percentage_Change), fill = Percentage_Change)) +
    geom_tile() +
    scale_fill_gradientn(colors = color_palette, limits = c(0, 100)) + 
    labs(title = paste(region_filter, "GHG Emission % Change (1990-2014)"),
         x = "Emission Category", y = "Country", fill = "% Change") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## **9️⃣ How European Countries' Emissions Have Evolved from 1990 to 2014**
```{r, fig.width=10, fig.height=6}
plot_heatmap("Europe")
```

## **🔟 Tracking Emission Changes in Non-European Countries (1990-2014)**
```{r, fig.width=10, fig.height=6}
plot_heatmap("Non-Europe")
```

